(Deep) Induction Rules for GADTs

### 2021-12-08 induction-paper/cpp2022/cpp2022-paper-final.tex
final cpp and final lfsa

\documentclass[sigplan,10pt]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}
\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
\usepackage{stmaryrd}
\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usepackage{bussproofs}
\EnableBpAbbreviations
\input{macros}

@@

@@

%\vspace*{0.05in}

@@

\section{Deep induction for ADTs and nested types}\label{sec:ADTs-and-nesteds}

@@

\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
      -0.2in 0in, fbox, center}

@@

\section{(Deep) induction for GADTs}\label{sec:deep-ind-GADTs}

@@

\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}

@@

\subsection{(Deep) induction for $\mathsf{Seq}$}\label{sec:ind-seq}

@@

\subsection{(Deep) induction for $\mathsf{LTerm}$}\label{sec:ind-lam}

@@

\[
\begin{array}{lll}
\mathsf{s} & : & \mathsf{String} \\
\mathsf{Q_A} & : & \mathsf{A \to Set} \\
\mathsf{Q_B} & : & \mathsf{B \to Set} \\
\mathsf{Q_C} & : & \mathsf{C \to Set} \\
\mathsf{T_A} & : & \mathsf{LType\, A} \\
\mathsf{T_B} & : & \mathsf{LType\, B} \\
\mathsf{t_B} & : & \mathsf{LTerm \, B} \\
\mathsf{t_C} & : & \mathsf{LTerm \, C} \\
\mathsf{t_{BA}} & : & \mathsf{LTerm \, (B \to A)} \\
\mathsf{ts} & : & \mathsf{List\, (LTerm\, B)}\\
\mathsf{liftA} & : & \mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A} \\
\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e} \\
\mathsf{lift_{T_B}} & : & \mathsf{LType^{\wedge} \, B\, Q_B\, T_B} \\
\mathsf{lift_{t_C}} & : & \mathsf{LTerm^{\wedge} \, C\, Q_C\, t_C} \\
\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}} \\
\mathsf{lift_{t_B}} & : & \mathsf{LTerm^{\wedge} \, B\, Q_B\, t_B} \\
\mathsf{liftE'} & : & \mathsf{Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e} \\
\mathsf{lift_{List}} & : & \mathsf{List^{\wedge} \, (LTerm\, B) \,
  (LTerm^{\wedge}\, B\, Q_B) \, ts}
\end{array}
\]
%$\mathsf{s : String}$, $\mathsf{Q_A
%  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
%  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B}$ $\mathsf{:
%  LType\,B}$, $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$,
%$\mathsf{t_{BA} : LTerm\,(B \to A)}$, $\mathsf{liftA :
%  LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
%  LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE :}$
%$\mathsf{Equal^{\wedge}}$ $\mathsf{A\, (B \to C)\, Q_A\, (Arr^{\wedge}
%  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
%  \, B\, Q_B\, T_B}$, $\mathsf{lift_{t_C}}$ $\mathsf{: LTerm^{\wedge}
%  \, C\, Q_C\, T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to
%  A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)}$ $\mathsf{t_{BA}}$,
%$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$,
%$\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A}$ and
%$\mathsf{(List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
%  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
Moreover, in the definition of $\mathsf{p_{ts}}$, \[\mathsf{PredMap :
  \forall\, (A : Set) \to (A \to Set) \to (A \to
  Set) \to Set }\] is the type constructor producing the type of
morphisms between predicates defined by\looseness=-1
\[\mathsf{PredMap \,A\, Q\,Q'\,
  = \forall\, (a : A) \to Q\,a \to Q'\,a}\] and
\[\begin{array}{l}
\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q' : A \to Set)
  \to}\\\
\hspace*{0.8in}\mathsf{PredMap\,A\,Q\,Q' \to}\\
\hspace*{1in}\mathsf{PredMap\,(List\,A)
  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}
\end{array}\]
which takes a morphism $\mathsf{f}$ of predicates and produces a
morphism of lifted predicates, is defined by
\[\mathsf{liftListMap\, A\, Q\, Q'\, m\, nil\, tt = tt}\]
(since $\mathsf{x : List^{\wedge}\,
  A\, Q\, nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$
of $\mathsf{\top}$), and by
\[\begin{array}{ll}
 & \mathsf{liftListMap\, A\, Q\, Q'\, m\,
  (cons\, a\, l')\, (y, x')}\\
= & \mathsf{(m\,a\,y, \,liftListMap\, A\,
  Q\, Q'\, m\, l'\, x')}
\end{array}\] (since $\mathsf{x : List^{\wedge}\, A\,
  Q\, (cons\, a\, l')}$ must be of the form $\mathsf{x = (y, x')}$
where $\mathsf{y : Q\,a}$ and $\mathsf{x' : List^{\wedge}\, A\, Q\,
  l'}$).

@@

\section{The general framework}\label{sec:framework}

@@

  $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We can define the first

@@

    K_\top\, t_1}$ and the function

@@

  a morphism of predicates between their liftings. The definition of
  $\mathsf{LTerm^\wedge EqualMap}$ is straightforwardly given by

@@

  constructor $\mathsf{Equal}$, together with the analogously defined

@@

 & \!\!\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
  ts)}\\
\quad\quad = & \!\!\mathsf{(K_\top , \, Equal^\wedge ListKT , \, L_{List^\wedge
    LTerm^\wedge KT} ) }

@@

>>>>

\documentclass[sigplan,screen]{acmart}
%\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}
\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
\usepackage{stmaryrd}
\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usepackage{bussproofs}
\EnableBpAbbreviations
\input{macros}

\copyrightyear{2022}
\acmYear{2022}
\setcopyright{acmlicensed}\acmConference[CPP '22]{Proceedings of the 11th
ACM SIGPLAN International Conference on Certified Programs and
Proofs}{January 17--18, 2022}{Philadelphia, PA, USA}
\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Conference
on Certified Programs and Proofs (CPP '22), January 17--18, 2022,
Philadelphia, PA, USA}
\acmPrice{15.00}
\acmDOI{10.1145/3497775.3503680}
\acmISBN{978-1-4503-9182-5/22/01}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752</concept_id>
<concept_desc>Theory of computation</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124</concept_id>
<concept_desc>Theory of computation~Semantics and reasoning</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010131.10010137</concept_id>
<concept_desc>Theory of computation~Categorical semantics</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation}
\ccsdesc[500]{Theory of computation~Semantics and reasoning}
\ccsdesc[500]{Theory of computation~Categorical semantics}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{GADTs, induction, proof assistants}

@@

@@

\section{Deep Induction for ADTs and Nested Types}\label{sec:ADTs-and-nesteds}

@@

\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
      -0.2in 0in, center}

@@

\section{(Deep) Induction for GADTs}\label{sec:deep-ind-GADTs}

@@

\subsection{(Deep) Induction for $\mathsf{Equal}$}\label{sec:ind-equal}

@@

\subsection{(Deep) Induction for $\mathsf{Seq}$}\label{sec:ind-seq}

@@

\subsection{(Deep) Induction for $\mathsf{LTerm}$}\label{sec:ind-lam}

@@

\[
\begin{array}{lll}
\mathsf{s} & : & \mathsf{String} \\
\mathsf{Q_A} & : & \mathsf{A \to Set} \\
\mathsf{Q_B} & : & \mathsf{B \to Set} \\
\mathsf{Q_C} & : & \mathsf{C \to Set} \\
\mathsf{T_A} & : & \mathsf{LType\, A} \\
\mathsf{T_B} & : & \mathsf{LType\, B} \\
\mathsf{t_B} & : & \mathsf{LTerm \, B} \\
\mathsf{t_C} & : & \mathsf{LTerm \, C} \\
\mathsf{t_{BA}} & : & \mathsf{LTerm \, (B \to A)} \\
\mathsf{ts} & : & \mathsf{List\, (LTerm\, B)}\\
\mathsf{liftA} & : & \mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A} \\
\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e} \\
\mathsf{lift_{T_B}} & : & \mathsf{LType^{\wedge} \, B\, Q_B\, T_B} \\
\mathsf{lift_{t_C}} & : & \mathsf{LTerm^{\wedge} \, C\, Q_C\, t_C} \\
\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}}
\end{array}\]

\[\begin{array}{lll}
\mathsf{lift_{t_B}} & : & \mathsf{LTerm^{\wedge} \, B\, Q_B\, t_B} \\
\mathsf{liftE'} & : & \mathsf{Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e} \\
\mathsf{lift_{List}} & : & \mathsf{List^{\wedge} \, (LTerm\, B) \,
  (LTerm^{\wedge}\, B\, Q_B) \, ts}
\end{array}
\]
%$\mathsf{s : String}$, $\mathsf{Q_A
%  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
%  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B}$ $\mathsf{:
%  LType\,B}$, $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$,
%$\mathsf{t_{BA} : LTerm\,(B \to A)}$, $\mathsf{liftA :
%  LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
%  LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE :}$
%$\mathsf{Equal^{\wedge}}$ $\mathsf{A\, (B \to C)\, Q_A\, (Arr^{\wedge}
%  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
%  \, B\, Q_B\, T_B}$, $\mathsf{lift_{t_C}}$ $\mathsf{: LTerm^{\wedge}
%  \, C\, Q_C\, T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to
%  A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)}$ $\mathsf{t_{BA}}$,
%$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$,
%$\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A}$ and
%$\mathsf{(List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
%  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
Moreover, in the definition of $\mathsf{p_{ts}}$, \[\mathsf{PredMap :
  \forall\, (A : Set) \to (A \to Set) \to (A \to
  Set) \to Set }\] is the type constructor producing the type of
morphisms between predicates defined by\looseness=-1
\[\mathsf{PredMap \,A\, Q\,Q'\,
  = \forall\, (a : A) \to Q\,a \to Q'\,a}\] and
\[\begin{array}{l}
\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q' : A \to Set)
  \to}\\\
\hspace*{0.8in}\mathsf{PredMap\,A\,Q\,Q' \to}\\
\hspace*{1in}\mathsf{PredMap\,(List\,A)
  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}
\end{array}\]
which takes a morphism $\mathsf{f}$ of predicates and produces a
morphism of lifted predicates, is defined by
\[\mathsf{liftListMap\, A\, Q\, Q'\, m\, nil\, tt = tt}\]
(since $\mathsf{x : List^{\wedge}\,
  A\, Q\, nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$
of $\mathsf{\top}$), and by
\[\begin{array}{ll}
 & \mathsf{liftListMap\, A\, Q\, Q'\, m\,
  (cons\, a\, l')\, (y, x')}\\
= & \mathsf{(m\,a\,y, \,liftListMap\, A\,
  Q\, Q'\, m\, l'\, x')}
\end{array}\] (since $\mathsf{x : List^{\wedge}\, A\,
  Q\, (cons\, a\, l')}$ must be of the form $\mathsf{x = (y, x')}$
where $\mathsf{y : Q\,a}$ and $\mathsf{x' : List^{\wedge}\, A\, Q\,
  l'}$).

@@

\section{The General Framework}\label{sec:framework}

@@

  $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We define the first

@@

    K_\top\, t_1}$ and the function\looseness=-1

@@

  a morphism of predicates between their liftings. We define
  $\mathsf{LTerm^\wedge EqualMap}$ straightforwardly by

@@

  constructor $\mathsf{Equal}$, together with the two analogously defined

@@

\pagebreak

\vspace*{-0.35in}

\[\begin{array}{ll}
 & \!\!\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
  ts)}\\
\quad\quad = & \!\!\mathsf{(K_\top , \, Equal^\wedge ListKT , \, L_{List^\wedge
    LTerm^\wedge KT} ) }
\end{array}\]
\end{itemize}

@@

\subsection*{Acknowledgments}

This work was supported by National Science Foundation award 1906388.
